---
title: "The Contrast Method for Literature-Based Discovery - Part 2: Short Reproducibility Study"
author: "Erwan Moreau"
date: "May 2021"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
library(rmarkdown)
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This is the documentation of the companion code for the paper **TODO**, which presents a novel approach for Literature-Based Discovery. The approach relies on contrasting different "views" of the data, the data being the biomedical literature (Medline and PubMed Central). The experiments use 8 relations which correspond to past discoveries in the domain of Neurodegenerative Diseases (NDs).

- The paper can be found at **TODO**
- The code is available at https://github.com/erwanm/lbd-contrast

For running these experiments two options are proposed: 

- The "full replication" offers more flexibility but requires much more computing resources (access to a cluster is recommended). For this option follow the instructions below then go to [Part 2](./short-reproducibility-study.html)
- The "short reproducibility study" proposes to use the exact same preprocessed dataset as described in the paper which can be downloaded from **TODO**. For this option you can go directly to [Part 2](./short-reproducibility-study.html).


## Data Requirements for the Short Reproducibility Study

For convenience the fully preprocessed dataset used in the paper is provided here **TODO**. The user should be able to reproduce the results presented in the paper by following the instructions below.


## Software Requirements

- This code requires R and a few R libraries: 
    - `reshape2`
    - `ggplot2`
    - `plyr`
    
The code can be loaded in the R interpreter with:

```{r}
source('contrast-experiments.R')
```

## Hardware Requirements

The [reproducibility study](#short-reproducility-study) only requires a recent enough computer with at least 8 GB RAM.


## Preparing for The Contrasts Experiments

### Loading the Datasets



```{r}
lbd_data<-read.table('lbd_data.tsv',header=TRUE,sep='\t',comment.char = "",quote="")
lbd_data_diso<-read.table('lbd_data.DISO.tsv',header=TRUE,sep='\t',comment.char = "",quote="")
```

### Calculating Association Measures

```{r}
lbd_data <- calculateAssociation(lbd_data)
lbd_data_diso <- calculateAssociation(lbd_data_diso)
```

### Specifying Methods Parameters

The parameters of the various methods are provided as a dataframe where each row contains the method id and the parameters. A few standard configuration methods can be generated with the `generateMethods()` function. For instance the following generates 12 methods using only PMI as association measure:

```{r}
methods0 <- generateMethods(measures='pmi')
```

The following will generate the same set of methods with the minimum frequency parameter set to either 1 or 3:

```{r}
generateMethods(measures='pmi',minFreqThresholds = c(1,3))
```

## Applying Methods

### Single Target



### Multiple Targets

